---
title: "G03 Apple Harvest"
author: "Norah Saarman"
date: "`r Sys.Date()`"
output:
  github_document:
    toc: true
editor_options: 
  markdown: 
    wrap: 72
---
  

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#Read in libraries
library(ggplot2)
library(ggpubr)
library(RColorBrewer)
library(plyr)
library(rstatix)
```

# Explore the data  

Read in data
```{r, read}
## Read in data set
apples <- read.csv(file="appleHarvest.csv")
#apples$city <- NULL #why this?

## Get length of harvest column
paste("N harvest data points =",length(apples$harvest))

```

Reformat, remove NAs, and structure the data:
```{r, remove NAs}
## Replace blanks "" with "blank"
apples$harvest[apples$harvest == ""] <- "blank"

## Revalue factor levels, "blank" = NA
apples$harvest <- revalue(apples$harvest, c("Minimal"="Minimal", "Normal"="Normal", "Above average"="Superior", "blank" = NA))

## omit NA
apples <- na.omit(apples)

## Retain categorical variables as factor
#confirm harvest is a factor
apples$harvest <- as.factor(apples$harvest) 
#confirm pruned is a factor
apples$pruned <- as.factor(apples$pruned) 

## Check that NAs omitted from harvest
apples$harvest

## Check that NAs omitted from purned status
apples$pruned

## Get length of final harvest column
paste("N harvest data points =",length(apples$harvest))

## Get length of final pruned column
paste("N pruning data points =",length(apples$pruned))
```
This now looks right (no blanks, no NA's), and the number of data points has been reduced from 241 to 144, and there is matching data for harvest and pruning status.


# ABSTRACT 

Fill in here.

# BACKGROUND

Fill in here.

# STUDY QUESTION & HYPOTHESIS

## Questions
Do pruned apple trees achieve higher categorical harvest ratings than unpruned trees?

## Hypothesis
We hypothesize that pruning improves yield for that year.

## Prediction
We predict that pruned trees will have a higher yield rating, with a rating more often marked as "Superior" or "Normal" than unpruned trees.

# METHODS & ANALYSIS  

## Method 1: Table of results, barplot visualization  

```{r table and barplot}
library(ggplot2)   # load library
library(scales)    # for percent_format()

## Table of results
table(apples$harvest,apples$pruned)

## Custom muted color palette (Minimal, Normal, Superior)
harvest_colors <- c(
  "Minimal"  = "#D4B86A",  # muted gold
  "Normal"   = "#97C29A",  # pastel green
  "Superior" = "#8CB7D6"   # pastel blue
)

## Stacked bar chart of yield categories by pruning status (ggplot2)
ggplot(apples, aes(x = pruned, fill = harvest)) +
  geom_bar(position = "fill") +
  scale_fill_manual(values = harvest_colors) +
  labs(
    x = "Pruning status",
    y = "Proportion of trees",
    fill = "Harvest"
  ) +
  scale_y_continuous(labels = percent_format()) +
  theme_minimal()
```
  
  
## Method 2: Chi-squared test

### Why Chi-square is appropriate

Both variables are categorical:  
- Pruned (No / Yes)  
- Harvest grade (Minimal / Normal / Superior)

Cells have sufficient expected counts (all > 5 except perhaps one, but overall ≥ 80% ≥ 5, which meets χ² assumptions).

We are testing whether the distribution of harvest grades differs by pruning status.

### Interpretation logic

Chi-square will tell you whether the grading categories occur in different proportions for pruned vs. unpruned trees.

Based on the table, the proportions look fairly similar (and “Superior” is rare in both), so we can expect non-significant results — aligning with the earlier ANOVA conclusion but now with the correct method.
```{r chi-squared}
# Contingency table
yield_tab <- table(apples$harvest, apples$pruned)

# Chi-square test of independence
chisq.test(yield_tab)
```
  
**Warning** appears because the superior counts are low and creates expected counts slightly below 5.

The rule is:

Chi-square approximation works well if ≥ 80% of cells have expected ≥ 5, and no cell is < 1. You have 1 small cell, but the rest are fine.
 
** So: Chi-square is usable, but you should confirm with Fisher’s Exact test** 
  
This is standard practice for categorical tests with a small cell or two.

```{r optional-fishers-exact}
fisher.test(yield_tab)
```
This confirms the result from the Chi-squared test.



# DISCUSSION

### Interpretation - Method 1 (Barplot)

The stacked bar plot shows that pruned and unpruned trees have very similar proportions of Minimal, Normal, and Superior harvest ratings. Although pruned trees appear to have slightly fewer Minimal yields and slightly more Normal yields, these differences are small and not visually convincing. Overall, the barplot suggests that pruning does not dramatically shift trees into higher or lower harvest categories in this dataset.

### Interpretation - Method 2 (Chi-squared test)

The chi-square test of independence formally tested whether pruning status was associated with harvest yield grade. The test was not significant (χ² = 2.37, df = 2, p = 0.31), meaning that the distribution of Minimal, Normal, and Superior yields did not differ between pruned and unpruned trees. This statistical result supports what is seen in the barplot: pruning did not lead to noticeably higher yield ratings in this dataset. Because the p-value was well above the common threshold of 0.05, we conclude that any differences observed are likely due to random variation rather than a real effect of pruning.
  
### Limitations  

Fill in here.

# CONCLUSION  

Fill in here.
  
# REFERENCES  
  
2025-26 apple production will reach nearly 279 million bushels. USApple.
(2025, August 15).
<https://usapple.org/news-resources/2025-26-apple-crop-outlook>

About the Wasatch Back Fruit Tree Project. Utah State University.
(n.d.). <https://extension.usu.edu/wasatch/WBFTP_about>

Kantor, L., & Blazejczyk, A. (2023, May 5). Apples and oranges are the
top U.S. fruit choices. Apples and oranges are the top U.S. fruit
choices \| Economic Research Service.
<https://www.ers.usda.gov/data-products/chart-gallery/chart-detail?chartId=58322>

University of Saskatchewan. (2018, February 2). Alternate or biennial bearing on fruit trees. College of Agriculture and bioresources. https://gardening.usask.ca/articles-and-lists/articles-disorders/disorder-alternate-or-biennial-bearing-on-fruit-trees.php  

Wikimedia Foundation. (2025, October 22). Dummy Variable (statistics). Wikipedia. https://en.wikipedia.org/wiki/Dummy_variable_(statistics) 
