---
title: "Warm-up mini-Report (Demo Report: “Grade My Work”): Mosquito Blood Hosts in \n Salt Lake City, Utah"
author: "Norah Saarman"
date: "`r Sys.Date()`"
output:
  github_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# ABSTRACT West Nile virus (WNV) circulates in bird–mosquito cycles with occasional human spillover. We asked whether house finch blood meals at surveillance sites predict the rate of WNV-positive mosquito pools in Salt Lake City. We summarized host composition by WNV presence with bar plots and tested whether house finch counts predict WNV positivity rate using a simple generalized linear model. Results are interpreted with key limitations and next steps for targeted surveillance.

# BACKGROUND

West Nile virus (WNV) is an arbovirus transmitted primarily by Culex mosquitoes, maintained through a bird–mosquito transmission cycle, and occasionally infecting humans and other mammals as incidental hosts. Understanding which bird species mosquitoes feed on helps identify potential amplifying hosts—those that develop sufficient viremia to transmit the virus back to mosquitoes.

Mosquito blood-meal analysis provides a direct link between vector and host. After mosquitoes are collected in the field, the abdomen is dissected and DNA is extracted from the blood meal. Host identity is determined by PCR amplification of a mitochondrial marker (cytochrome oxidase I) followed by MinION nanopore sequencing and comparison to a reference database such as GenBank using BLASTN. These methods allow us to connect mosquito host choice with virus prevalence in local mosquito pools.

Previous studies (Komar et al. 2003) demonstrated that some bird species, including house finches and American robins, maintain viremia long enough to infect mosquitoes efficiently. Therefore, we hypothesize that sites where mosquitoes have fed more frequently on these species will also show higher rates of WNV-positive mosquito pools. Figure 1 illustrates the variation in viremia duration among North American birds from Komar et al. (2003).

```{r background, include=TRUE, eval=TRUE, warning=FALSE, fig.align='left', fig.width=4.75, fig.height=7}
# Manually transcribe duration (mean, lo, hi) from the last table column
duration <- data.frame(
  Bird = c("Canada Goose","Mallard", 
           "American Kestrel","Northern Bobwhite",
           "Japanese Quail","Ring-necked Pheasant",
           "American Coot","Killdeer",
           "Ring-billed Gull","Mourning Dove",
           "Rock Dove","Monk Parakeet",
           "Budgerigar","Great Horned Owl",
           "Northern Flicker","Blue Jay",
           "Black-billed Magpie","American Crow",
           "Fish Crow","American Robin",
           "European Starling","Red-winged Blackbird",
           "Common Grackle","House Finch","House Sparrow"),
  mean = c(4.0,4.0,4.5,4.0,1.3,3.7,4.0,4.5,5.5,3.7,3.2,2.7,1.7,6.0,4.0,
           4.0,5.0,3.8,5.0,4.5,3.2,3.0,3.3,6.0,4.5),
  lo   = c(3,4,4,3,0,3,4,4,4,3,3,1,0,6,3,
           3,5,3,4,4,3,3,3,5,2),
  hi   = c(5,4,5,5,4,4,4,5,7,4,4,4,4,6,5,
           5,5,5,7,5,4,3,4,7,6)
)

# Choose some colors
cols <- c(rainbow(30)[c(10:29,1:5)])  # rainbow colors

# horizontal barplot
par(mar=c(5,12,2,2))  # wider left margin for names
bp <- barplot(duration$mean, horiz=TRUE, names.arg=duration$Bird,
              las=1, col=cols, xlab="Days of detectable viremia", xlim=c(0,7))

# add error bars
arrows(duration$lo, bp, duration$hi, bp,
       angle=90, code=3, length=0.05, col="black", xpd=TRUE)
```

# STUDY QUESTION and HYPOTHESIS

## Question(s)  
Which bird species acts as an amplifying host of WNV in Salt Lake City? Do house finch blood meals predict site-level WNV positivity?

## Hypothesis 
House finches are acting as important amplifying hosts of WNV in Salt Lake City.
  
## Prediction(s)
Sites with more house finch blood meals have higher WNV positivity rates. Therefore, I predict that house finch blood-meal counts will be positively associated with WNV-positive pool rate.

# MOTHODS AND RESULTS

Analysis 1: Bar plots of host composition by WNV status To evaluate whether host community composition differed between sites with and without WNV detections, we summarized blood-meal counts for each identified host species according to whether the site had one or more WNV-positive pools. 

```{r barplot, warning=FALSE}
## import counts_matrix: data.frame with column 'loc_positives' (0/1) and host columns 'host_*'
counts_matrix <- read.csv("./bloodmeal_plusWNV_for_BIOL3070.csv")

## 1) Identify host columns
host_cols <- grep("^host_", names(counts_matrix), value = TRUE)

if (length(host_cols) == 0) {
  stop("No columns matching '^host_' were found in counts_matrix.")
}

## 2) Ensure loc_positives is present and has both levels 0 and 1 where possible
counts_matrix$loc_positives <- factor(counts_matrix$loc_positives, levels = c(0, 1))

## 3) Aggregate host counts by loc_positives
agg <- stats::aggregate(
  counts_matrix[, host_cols, drop = FALSE],
  by = list(loc_positives = counts_matrix$loc_positives),
  FUN = function(x) sum(as.numeric(x), na.rm = TRUE)
)

## make sure both rows exist; if one is missing, add a zero row
need_levels <- setdiff(levels(counts_matrix$loc_positives), as.character(agg$loc_positives))
if (length(need_levels)) {
  zero_row <- as.list(rep(0, length(host_cols)))
  names(zero_row) <- host_cols
  for (lv in need_levels) {
    agg <- rbind(agg, c(lv, zero_row))
  }
  ## restore proper type
  agg$loc_positives <- factor(agg$loc_positives, levels = c("0","1"))
  ## coerce numeric host cols (they may have become character after rbind)
  for (hc in host_cols) agg[[hc]] <- as.numeric(agg[[hc]])
  agg <- agg[order(agg$loc_positives), , drop = FALSE]
}

## 4) Decide species order (overall abundance, descending)
overall <- colSums(agg[, host_cols, drop = FALSE], na.rm = TRUE)
host_order <- names(sort(overall, decreasing = TRUE))
species_labels <- rev(sub("^host_", "", host_order))  # nicer labels

## 5) Build count vectors for each panel in the SAME order
counts0 <- rev(as.numeric(agg[agg$loc_positives == 0, host_order, drop = TRUE]))
counts1 <- rev(as.numeric(agg[agg$loc_positives == 1, host_order, drop = TRUE]))

## 6) Colors: reuse your existing 'cols' if it exists and is long enough; otherwise generate
if (exists("cols") && length(cols) >= length(host_order)) {
  species_colors <- setNames(cols[seq_along(host_order)], species_labels)
} else {
  species_colors <- setNames(rainbow(length(host_order) + 10)[seq_along(host_order)], species_labels)
}

## 7) Shared x-limit for comparability
xmax <- max(c(counts0, counts1), na.rm = TRUE)
xmax <- if (is.finite(xmax)) xmax else 1
xlim_use <- c(0, xmax * 1.08)

## 8) Plot: two horizontal barplots with identical order and colors
op <- par(mfrow = c(1, 2),
          mar = c(4, 12, 3, 2),  # big left margin for species names
          xaxs = "i")           # a bit tighter axis padding

## Panel A: No WNV detected (loc_positives = 0)
barplot(height = counts0,
        names.arg = species_labels, 
        cex.names = .5,
        cex.axis = .5,
        col = rev(unname(species_colors[species_labels])),
        horiz = TRUE,
        las = 1,
        xlab = "Bloodmeal counts",
        main = "-WNV Sites",
        xlim = xlim_use)

## Panel B: WNV detected (loc_positives = 1)
barplot(height = counts1,
        names.arg = species_labels, 
        cex.names = .5,
        cex.axis = .5,
        col = rev(unname(species_colors[species_labels])),
        horiz = TRUE,
        las = 1,
        xlab = "Bloodmeal counts",
        main = "+WNV Sites",
        xlim = xlim_use)

par(op)

## Keep the colors mapping for reuse elsewhere
host_species_colors <- species_colors
```
  
## Analysis 2: Relationship between house finch blood meals and WNV positivity rate 

To formally test whether mosquito feeding on house finches was associated with local WNV transmission, we fit a simple linear model with the proportion of WNV-positive pools per site (loc_rate) as the response variable and the number of house finch blood meals (host_House_finch) as the predictor. This model quantifies the strength and direction of the association observed in the descriptive plots. Model assumptions (normality and homoscedasticity of residuals) were assessed visually. We report the estimated regression coefficient (β) and associated p-value as indicators of statistical support.

```{r second-analysis-or-plot}

#glm: house-finch correlated with WNV positivity rate of pools?
glm2 <- glm(loc_rate ~ host_House_finch,
            data = counts_matrix)
#summary(glm2)
```




DISCUSSION

Results from this analysis supports the hypothesis that mosquito feeding on house finches is positively associated with WNV detection in Salt Lake City. While these relationships are correlative, they highlight how integrating molecular host identification with infection data can reveal ecological drivers of virus amplification. Further work incorporating multi-year sampling and environmental covariates (e.g., mosquito abundance, habitat type) would strengthen inference and guide targeted vector control.
  
Analysis 1: Host composition by WNV presence

Host identity was determined using DNA extraction and sequencing of the vertebrate cytochrome oxidase I gene, followed by BLAST comparison to the GenBank reference database. We aggregated counts by site-level WNV status (loc_positives: 0 or 1) and visualized the results as paired horizontal bar plots. This visualization highlights which hosts are most frequently represented at WNV-positive sites and serves as a qualitative indicator of potential amplifying species.
  
Analysis 2: Relationship between house finch blood meals and WNV positivity rate

This analysis and the plot show that the regression coefficient for house finch blood meals was positive (β ≈ 0.03 ± 0.006, p = 4.54e-05), indicating a highly statistically significant increase in WNV positivity rate with higher house finch representation. This supports the hypothesis that feeding on house finches contributes to local virus amplification.
REFERENCES     

2. ChatGPT. OpenAI, version Jan 2025. Used as a reference for functions such as plot() and to correct syntax errors. Accessed `r Sys.Date()`.



