---
title: "Group 6 Help (simulated data)"
author: "Norah Saarman"
date: "`r Sys.Date()`"
output:
  github_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# ABSTRACT


# BACKGROUND

```{r background, include=TRUE, eval=TRUE}
#background
```

# STUDY QUESTION and HYPOTHESIS

## Question(s)  


## Hypothesis 

  
## Prediction(s)


# METHODS AND RESULTS

Data simulated and given to us by instructor, based on https://doi.org/10.1158/1078-0432.CCR-10-3028... The original paper [did what?]...


## Analysis 1: Visualize with a Boxplot

First, we visualized the impact of RHBDF2 genotype on normalized pain relief...

```{r boxplot}
library(ggplot2)
library(dplyr)

# Read in simulated data
dat <- read.csv("epos_style_rs12948783_simulated_large.csv")

dat$carrier <- factor(dat$carrier, levels = c("AA_or_GA","GG"))

means <- dat %>%
  group_by(carrier) %>%
  summarise(m = mean(norm_pain_relief), .groups = "drop")

p2 <- ggplot(dat, aes(carrier, norm_pain_relief)) +
  geom_boxplot(width = 0.55, outlier.shape = NA) +
  geom_jitter(width = 0.15, alpha = 0.15, size = 0.8) +
  stat_summary(fun = mean, geom = "point", size = 3) +
  stat_summary(fun.data = mean_se, geom = "errorbar", width = 0.15) +
  geom_segment(data = data.frame(x = 1, xend = 2,
                                 y = means$m[means$carrier=="AA_or_GA"],
                                 yend = means$m[means$carrier=="GG"]),
               aes(x = x, xend = xend, y = y, yend = yend)) +
  coord_cartesian(ylim = c(80, 110)) +
  labs(x = "rs12948783 (AA/GA vs GG)",
       y = "Normalized pain relief (%)") +
  theme_classic(base_size = 13)

p2
# ggsave("boxplot_collapsed.png", p2, width = 5, height = 4, dpi = 300)
```
Next, we performed an ANOVA to test if pain relief was statistically different between the different RHBDF2 genotypes.

```{r anova, warning=FALSE}

# One-way ANOVA (3 genotypes):
dat$genotype <- factor(dat$genotype, levels = c("AA","GA","GG"))
summary(aov(norm_pain_relief ~ genotype, data = dat))
TukeyHSD(aov(norm_pain_relief ~ genotype, data = dat))


#Two-group ANOVA (AA/GA vs GG), matching the figure:
dat$carrier <- factor(dat$carrier, levels = c("AA_or_GA","GG"))
summary(aov(norm_pain_relief ~ carrier, data = dat))

# 2-way ANOVA with sex as a covariate:
summary(aov(norm_pain_relief ~ genotype + sex, data = dat))
```  
The results show...


## Analysis 2: 


# DISCUSSION

Results from this analysis...
  
## Analysis 1: Boxplot and ANOVA

  
## Analysis 2: 



# CONCLUSION


# REFERENCES     

1. Galvan, A., Skorpen, F., Klepstad, P., Knudsen, A. K., Fladvad, T., Falvella, F. S., Pigni, A., Brunelli, C., Caraceni, A., Kaasa, S., & Dragani, T. A. (2011). Multiple loci modulate opioid therapy response for cancer pain. Clinical Cancer Research, 17(13), 4581–4587. https://doi.org/10.1158/1078-0432.CCR-10-3028 

2. ChatGPT. OpenAI, version Jan 2025. Used as a reference for... and to correct syntax errors... and... Accessed `r Sys.Date()`.



